---
title: "Social Sequence Analysis"
output: html_document
author: Xuewen Yan
date: "2025-02-26"
editor_options: 
  chunk_output_type: console
---

This file draws extensively on Emanuela Struffolino and Anette Fasang's workshop on sequence analysis  https://pdhp.isr.umich.edu/workshops/sequence-analysis-for-social-science/

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data preparation

### Install and load packages
```{r , echo=TRUE, eval=FALSE}
# Uncomment to install

# install.packages("tidyverse")       
# install.packages("haven")            
# install.packages("TraMineR")         
# install.packages("TraMineRextras")   
# install.packages("WeightedCluster")

library("tidyverse")       ### most important packages from the tidyverse
library("haven")            ### read data stored in various formats
library("TraMineR")         ### "Life Trajectory Miner for R (and was inspired by the authors' taste for Gewürztraminer wine)."
library("TraMineRextras")   ### TraMineR Extensions
library(WeightedCluster)
```

```{r echo=FALSE,eval=TRUE}
suppressPackageStartupMessages(library("tidyverse"))       ### most important packages from the tidyverse
suppressPackageStartupMessages(library("haven"))            ### read data stored in various formats
suppressPackageStartupMessages(library("TraMineR"))         ### "Life Trajectory Miner for R (and was inspired by the authors' taste for Gewürztraminer wine)."
suppressPackageStartupMessages(library("TraMineRextras"))   ### TraMineR Extensions
suppressPackageStartupMessages(library(WeightedCluster))
```

### Import dta & label factors
```{r }
family <- read_dta("PartnerBirthbio.dta") # make sure the dta file is in the same folder/directory as the R code
names(family) %>% tail() # tail(names(family))

family <- as_factor(family,only_labelled = T) 
longlab <- levels(family$state1)

shortlab <- c("S", "Sc", "LAT", "LATc",
                     "COH", "COHc",
                     "MAR", "MARc1", "MARc2+")

# levels(family$state1) <- shortlab
family <- family %>% 
  mutate(across(starts_with("state"), ~ factor(.x, levels = levels(.x), labels = shortlab)))

```

## Seq object creation

```{r}
a <- which(names(family) == "state1");a
names(family) %>% tail()
b <- which(names(family) == "state264");b

family_month_seq <- seqdef(family[a:b],
                           alphabet = levels(family$state1),
                           states = shortlab,
                           labels = longlab,
                           # cpal = colspace1, 
                           weights = family$weight40, # use weights
                           xtstep = 24) # specifies interval of tick-marks
help(seqdef)


# Optional: change granularity --> years instead of months (using modal values)

family_year_seq <- seqgranularity(family_month_seq,
                                  tspan=12, 
                                  method="mostfreq")

# Optional: Spell long-form convertion
family_spell <- seqformat(family[a:b], from = "STS", to = "SPELL")
family_sts <- seqformat(family_spell, from = "SPELL", to = "STS",
  id = "id", begin = "begin", end = "end", status = "states",
  process = TRUE, limit = b) #set limit to the largest row number by id
```

## Seq descriptives

### Summary stats
```{r }
# variance and sd of the total time spent in the states, and se of the mean
seqmeant(family_month_seq, serr = TRUE)

# same calculation, by gender & export 
family$sex %>% levels()
t1 <- seqmeant(family_month_seq[family$sex == "Male",])
t2 <- seqmeant(family_month_seq[family$sex == "Female",])
ob1 <- data.frame(t1,t2);
ob1
write.csv(ob1,file = "mytb.csv")

# proportions of time spent in each state for the sample
seqmeant(family_month_seq, prop = TRUE)

# average number of episodes/spells (stability) for the sample
seqmeant(seqdss(family_month_seq),serr = TRUE)
```

### Individual-level stats
```{r}
# Number of transitions for each participant
family$n_trans <- seqtransn(family_month_seq); head(family$n_trans,5)
mean(family$n_trans)

# Total time in each state for each participant
time_in_state <- seqistatd(family_month_seq) %>% as_tibble() #returns a matrix col = each state; row = individual; cell = # of time points in each state; as_tibble converts this to df
names(time_in_state) <- paste0("dur_",names(time_in_state))
head(time_in_state)
family <- tibble(family,as_tibble(time_in_state))

# Number of episodes/spells in each state for each participant
n_spells <- seqistatd(seqdss(family_month_seq)) %>% as_tibble()
names(n_spells) <- paste0("nspell_",names(n_spells))
family <- tibble(family, as_tibble(n_spells))

# check data
names(family) %>% tail(20)
```

### Transition matrix (omitted, explore yourself)
```{r }
# ... using yearly sequence data 
transmat_year <- seqtrate(family_year_seq)
round(transmat_year,2)


# ... using yearly sequence data converted to DSS format (spell perspective)
transmat_dss <- seqtrate(seqdss(family_year_seq))
round(transmat_dss,2)

```

### Modal sequence summary & visuals
```{r }
tp <- seqmodst(family_year_seq);tp # again, can do similar tricks for groups
print(tp, format = "SPS")

# visualize modal sequence
seqmsplot(family_year_seq, 
          group = family$sex, 
          border=NA,
          ylab = "Relative frequency",
          xtlab = seq(18,44),
          # xtstep = 2, 
          xlab = "Age",
          main = NULL)
# locator(1)
rect(xleft = -2, ybottom = 1.2, xright = max(seq_len(ncol(family_year_seq))), ytop = 1, 
     col = "white", border = NA)
mtext("Male", side = 3, line = -1, adj = 0.1, cex = 1.2)
mtext("Female", side = 3, line = -1, adj = 0.7, cex = 1.2)
```


## Seq descriptive plots

### State distribution plot
```{r }
# Plotting the general distribution of states in the population
seqdplot(family_year_seq, 
         main = "Yearly Granularity",
         # group = family$sex,
         border = NA,
         xlab = "Age",
         xtlab = seq(18,44,by = 1))
```

###  Index plot
```{r}
# plotting individual sequences, only feasible for a limited number of cases)
seqIplot(family_year_seq, 
         # group = family$sex,
         main = "Yearly Granularity",
         border = NA)
```

### Weighted index plot
```{r }
# plotting weighted individual sequences, selects a subset of sequences proportionally based on state frequencies )
seqiplot(family_year_seq, 
         group = family$sex,
         main = "Yearly Granularity",
         border = NA)
```

```{r Relative frequency index plot (omitted, explore yourself) }
# Relative freq index plot balances the summary structure of state distribution plotting and index plotting by calculating some representative sequences and then individually plotting them. (Fasang & Liao 2014) Essentially, only the medoid sequences are plotted. A medoid sequence is a sequence that has minimal dissimilarity from other sequences that are classified in the same cluster. This quietly conducts some clustering analysis.

? seqdist
family_year_om <- seqdist(family_year_seq, 
                          method="OM", sm= "CONSTANT")

seqplot.rf(family_year_seq,
           diss=family_year_om,
           k=30)
```

## Seq comparison and clustering

sm: substiution cost matrix, "CONSTANT" assigns a constant value of any transitions between states; "TRATE" does data driven measure where less common transitions are more expensive. indel by default =  "auto" = max(sm)/2;

### Optimal matching
```{r}
family_year_om <- seqdist(family_year_seq, 
                          method="OM", sm= "CONSTANT", 
                          indel = "auto") 

# much higher indel costs
om_highcost <-seqdist(family_year_seq, 
              method = "OM", 
              indel = 4, 
              sm = "CONSTANT",
              norm = "auto")

# manually specifying sm based on theory: (omitted, explore yourself)
   # substitution=1 is one if different number of children but same partnership status 
   # substitution=2 any substitution of different partnership statuses
summary(family_year_seq)
theo <- matrix(
  c(0,1,2,2,2,2,2,2,2,
    1,0,2,2,2,2,2,2,2,
    2,2,0,1,2,2,2,2,2,
    2,2,1,0,2,2,2,2,2,
    2,2,2,2,0,1,2,2,2,
    2,2,2,2,1,0,2,2,2,
    2,2,2,2,2,2,0,1,1,
    2,2,2,2,2,2,1,0,1,
    2,2,2,2,2,2,1,1,0),
  nrow = 9, ncol = 9, byrow = TRUE)

om_theo<-seqdist(family_year_seq, 
                 method = "OM", 
                 indel = 1,
                 sm = theo,
                 norm = "auto")

# Should compare clustering results based on different cost specification strategies

```

### Hierarchical clustering
```{r }
# We use ward distance that minimizes residual variance. 
# The goal of clustering is to group similar sequences together while making sure that sequences in different clusters are as dissimilar as possible. 
class(family_year_om)
class(as.dist(family_year_om))

fam.ward1 <- hclust(as.dist(family_year_om), 
                    method = "ward.D", 
                    members = family$weight40)
class(fam.ward1)
plot(fam.ward1)

```

### Evaluate clustering quality
AWS: average silhouette width; average coherence of the assigned sequences across the clusters, larger = better
```{r }
fam.ward.test <- as.clustrange(fam.ward1, 
                               diss = family_year_om,
                               weights =family$weight40, 
                               ncluster = 10) 


fam.ward.test
```

###  Extract clustering solutions
```{r}
family$cl5 <-cutree(fam.ward1 , k = 5)
family$cl4 <-cutree(fam.ward1, k = 4)

# Compare 4- and 5-cluster solutions
table(family$cl5, family$cl4)
```

### Visualize sequence clusters 
```{r}
seqdplot(family_year_seq, 
         group = family$cl4, 
         border = NA)

seqdplot(family_year_seq, 
         group = family$cl4, 
         border = FALSE)

seqdplot(family_year_seq, 
         group = family$cl5, 
         border = FALSE)

seqiplot(family_year_seq, 
         group = family$cl4, 
         border = FALSE)

seqiplot(family_year_seq, 
         group = family$cl5, 
         border = NA)

seqIplot(family_year_seq, 
         group = family$cl4, 
         border = NA)
```

## Regression analysis
```{r}
# We already have the clustering groups stored! We also created several variables: n_trans; and the # of time points in a state for each state 
getwd()
family$dur_MARc2orplus<- family$`dur_MARc2+` # name MARc2+ is illegal for write_dta
family$nspell_MARc2orplus<- family$`nspell_MARc2+` # name MARc2+ is illegal for write_dta
family$`dur_MARc2+` <- NULL #delete a column
family$`nspell_MARc2+` <- NULL #delete a column

write_dta(family, path = "family_cl.dta") 
```

## Colors!
```{r}
# install.packages("RColorBrewer")
# install.packages("colorspace")

library(RColorBrewer)
library(colorspace)
# Inspect predefined color palettes

display.brewer.pal(3, "Blues")
display.brewer.pal(7, "Blues")

# Choosing colors

col1 <- brewer.pal(3, "Blues")[2:3]   # takes the 2nd to 3rd types of brewer blue to color 1 (eventually use this for Single states)
col2 <- brewer.pal(3, "Greens")[2:3]  # LAT states
col3 <- brewer.pal(3, "Oranges")[2:3] # Cohabitation states
col4 <- brewer.pal(7, "Purples")[5:7] # Marriage states

# Define complete color palette

colspace1 <- c(col1, col2, col3, col4) # can specify color pallette option in seqdef!

# Inspect color palette

swatchplot(colspace1)
```


```{r echo=TRUE,eval=TRUE}
# seq def again with self-defined coloring
family_month_seq2 <- seqdef(family[a:b],
                           alphabet = levels(family$state1),
                           states = shortlab,
                           labels = longlab,
                           cpal = colspace1, 
                           weights = family$weight40, # use weights
                           xtstep = 24) # specifies interval of tick-marks
family_year_seq2 <- seqgranularity(family_month_seq2,
                                  tspan=12, 
                                  method="mostfreq")

seqiplot(family_year_seq2,
         group = family$sex,
         border = NA,
         #ylab = "Weighted frequencies",
         xtlab = seq(18,44),
         xlab = "Age")
```

